\documentclass{article}

% Camera-ready: \usepackage[final]{neurips_2024}
% Preprint:     \usepackage[preprint]{neurips_2024}
\usepackage[preprint]{neurips_2024}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts, amsmath, amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\title{The Pareto Landscape of Moral Reasoning}

\author{%
  Pierre Beckmann\\
  University of Bern\\
  \texttt{pierrebeckmann@gmail.com}
}

\begin{document}

\maketitle

\begin{abstract}
Evaluating moral reasoning is hard because the moral domain lacks the ground
truth that organises evaluation elsewhere: competent reasoners can start from
similar intuitions and reach different defensible conclusions. We argue that
this is not a measurement failure but a structural feature, and we propose a
conceptual framework that takes it seriously. Moral reasoning, on our account,
aims at a \emph{good epistemic state}: a pair of commitments and principles
that hangs together. Such a state is graded along two axis families that
genuinely conflict: epistemic desiderata (account, systematicity, faithfulness)
and value loadings (freedom, well-being, justice, dignity, solidarity). The
rational object of study is then the \emph{Pareto frontier} of attainable
epistemic states. We develop a vocabulary for the shapes this frontier takes
(convex segments, concave segments, knees, plateaus, corner attractors,
dominated regions), and we recast familiar reasoning methods, reflective
equilibrium foremost, as strategies for navigating the
landscape toward the frontier. We sketch the corpus that would let this
framework be studied empirically, and outline what such study would reveal
about moral reasoning in humans and in language models.
\end{abstract}


\section{Motivation}
\label{sec:motivation}

Current alignment work focuses on the moral \emph{content} a model holds:
constitutional principles, character training, value specifications. Less
attention has gone to moral reasoning as a \emph{capacity}~\citep{carlsmith2026},
and existing evaluations tend to conflate the two \citep{haas2026}. But a
value-aligned model whose reasoning fails when its values come into tension
is still unsafe: it can produce outputs that look like principled moral
thought without being so. Diagnosing such failures is hard because the moral
domain lacks a ground truth against which to score answers \citep{queloz2025}:
competent reasoners can reach different conclusions without either being
simply wrong.

This paper proposes a framework that takes the absence of ground truth as a
\emph{structural feature} rather than a measurement failure. We articulate
what moral reasoning aims at, what makes its outputs better or worse without
a single right answer, and how the multi-objective optimisation literature
\citep{deb2001multiobjective} gives us a precise vocabulary for the kinds of
trade-offs the moral domain actually exhibits. Familiar methods such as
reflective equilibrium \citep{rawls1971,knight2023}, recently formalised by
\citet{beisbart2021} and studied at simulation scale by \citet{freivogel2023},
reappear as \emph{strategies for navigating} this landscape, comparable in
principle to alternative strategies and amenable to empirical study once a
suitable corpus is in place.


\section{Framework}
\label{sec:framework}

We develop the framework in four steps: introducing the conceptual objects
moral reasoning operates on, the axes along which their combinations are
graded, the Pareto frontier those axes define, and a vocabulary for the
shapes the frontier can take.

\subsection{Commitments, principles, epistemic state}
\label{sec:concept}

An \textbf{ethical situation} confronts an agent with a decision: several
actions are open, and none is clearly right or wrong. On encountering such a
situation the agent has \textbf{commitments}: pre-theoretical verdicts about
how to act in the case, of the form ``in situation $s$, action $a$ is
required / permissible / forbidden.'' Across the body of situations the
agent has considered, these commitments form a set $C$. A reasoner holding
$C$ alone has no resources to handle a new situation, to defend a verdict
against an objection, or to notice when two commitments pull in opposite
directions.

This is where \textbf{principles} enter. A principle $p$ is a general moral
rule, independent of any particular situation, that an agent endorses as
justifying some of their commitments. The agent's principles together form
a set $P$. Principles do three kinds of work simultaneously: they
\emph{account} for commitments (giving each commitment a reason), they
\emph{generalise} to situations the agent has not yet considered (a single
principle yields verdicts across many cases), and they \emph{regulate} the
commitments (revealing inconsistency when two commitments cannot both
follow from the same principle).

The output of moral reasoning is an \textbf{epistemic state}
$\sigma = (C, P)$: a set of commitments paired with a set of principles
that hangs together. We take the goal of moral reasoning to be reaching a
\emph{good} epistemic state. Two families of demand pull on what
\emph{good} means here. The state should satisfy the agent's
\emph{epistemic} expectations of reasoning: the principles should account
for the commitments, they should be simple and broadly applicable rather
than a hodgepodge of one-offs, and the commitments should not drift
arbitrarily from the agent's starting intuitions $C_0$. The state should
also fit the agent's \emph{ethical} profile: the principles should express
the values and deeper ethical frameworks the agent endorses on reflection,
and where values trade off, the trade-off the state strikes should be one
the agent can stand behind. We make these two families precise in
Section~\ref{sec:good}.



\subsection{What makes an epistemic state good}
\label{sec:good}

Two axis families grade $\sigma$ simultaneously. They genuinely conflict, in
the sense that improving one typically costs another.

\paragraph{Epistemic axes.} Properties of $\sigma$ as reasoning. Following
the formal account of \citet{beisbart2021}, we work with three; others
could be envisaged (simplicity, scope, robustness) and the framework
accommodates them without modification.
\begin{itemize}[leftmargin=*, itemsep=2pt]
  \item \textsc{Account}: how much the principles in $P$ entail or support
    the commitments in $C$, penalising contradictions, gaps, and surplus
    content.
  \item \textsc{Systematicity}: how simple, unifying, and broadly applicable
    $P$ is, as opposed to a hodgepodge of case-by-case stipulations.
  \item \textsc{Faithfulness}: how well the current commitments respect the
    agent's initial intuitions $C_0$, penalising arbitrary drift.
\end{itemize}

\paragraph{Ethical axes.} Properties of $\sigma$ as a moral stance. Each
principle $p \in P$ carries a fuzzy \emph{value loading} $\boldsymbol{v}(p)
\in [0, 1]^K$ over a small set of basic moral values. We work with five:
\textsc{freedom}, \textsc{well-being}, \textsc{justice}, \textsc{dignity},
and \textsc{solidarity}, chosen so that the central tensions of moral life
(liberty vs.\ community, fairness vs.\ welfare, welfare vs.\ dignity) appear
directly as trade-offs among axes. The state $\sigma$ inherits an aggregate
value position by aggregating its principles' loadings: the more a value is
activated by the principles in $P$, the more the state weighs on that axis.
Concretely, $V(\sigma) = \tfrac{1}{|P|} \sum_{p \in P} \boldsymbol{v}(p)
\in [0, 1]^K$, so locating an epistemic state in the ethical-value subspace
amounts to checking which values each of its principles activates.

\paragraph{Two families, two axes of disagreement.} The epistemic family is
broadly agent-independent: every reasoner has some stake in coherence. The
ethical family is agent-dependent: different agents weight different values,
and this pluralism is a constitutive feature of the moral domain rather than
noise to be averaged out.

\subsection{Reasoning moves}
\label{sec:moves}

Reaching a good $\sigma$ involves recognisable operations on $C$ and $P$.
We propose a candidate list of typed primitives, each with an
input-output signature that makes it addressable as a unit of evaluation
independent of the strategy that composes it. The first four are needed,
to varying degrees, by any moral reasoning strategy; the last two are
characteristic of \emph{dialectical} strategies and will reappear in
Section~\ref{sec:strategies} as the central moves of reflective
equilibrium.
\begin{itemize}[leftmargin=*, itemsep=2pt]
  \item \textbf{Property elicitation} \;$s \to \{\varphi_1, \dots, \varphi_m\}$.\,
    From a situation, make explicit the ethically relevant properties: who
    is affected, what interests are at stake, the kind of action on offer.
    The basis on which a commitment will be formed.
  \item \textbf{Principle proposal} (abduction) \;$\{c_1, \dots, c_n\} \to p$.\,
    From a set of commitments, propose a candidate principle that justifies
    them. The inductive move from cases to rule.
  \item \textbf{Principle deduction} \;$(p, s) \to c$.\, From a candidate
    principle and a situation, derive the commitment the principle implies.
    The downward complement of proposal.
  \item \textbf{Coherence checking} \;$\sigma \to \mathbb{R}^{3+K}$.\, Score
    the current state on the epistemic and ethical axes (Section~\ref{sec:good}).
    A meta-operation: it does not move $\sigma$, but it tells any other
    operation whether its candidate move improves things.
  \item \textbf{Commitment revision} \;$(c, p) \to c'$.\, Adjust a
    commitment in light of a principle that conflicts with it.
  \item \textbf{Principle revision} \;$(p, c) \to p'$.\, Adjust a principle
    in light of a commitment that conflicts with it.
\end{itemize}
The dialectical alternation of the two revisions \citep{rawls1971,beisbart2021}
is what gives reflective equilibrium its name; we make the strategy
explicit in Section~\ref{sec:strategies}, after spelling out the space the
moves take place in.


\subsection{The Pareto frontier and its projections}
\label{sec:space}

We can now state the framework precisely. Let $\Sigma$ be the set of all
internally consistent epistemic states. Each $\sigma \in \Sigma$ has a
position in the joint objective space
\[
  \bigl(A(\sigma),\, S(\sigma),\, F(\sigma \mid C_0),\, V_1(\sigma), \dots, V_K(\sigma)\bigr)
  \in [0,1]^{3+K}.
\]
Multi-objective optimisation~\citep{deb2001multiobjective} gives us the right
vocabulary for the structure of this space when its axes genuinely conflict.

\paragraph{Pareto dominance.} A state $\sigma'$ \emph{Pareto dominates}
$\sigma$, written $\sigma \prec \sigma'$, if every objective at $\sigma'$ is
at least as large as at $\sigma$ and at least one is strictly larger.

\paragraph{The Pareto frontier.} The frontier $\Pi$ is the set of
non-dominated states in $\Sigma$:
\[
  \Pi \;=\; \{\,\sigma \in \Sigma : \nexists\, \sigma' \in \Sigma
                \text{ with } \sigma \prec \sigma'\,\}.
\]
$\Pi$ is what we mean by \emph{the space of defensible epistemic states}.
For any state off the frontier, some defensible move strictly improves it
on every axis the agent cares about; for any state on the frontier, no such
move exists.

\paragraph{Untangling $\Pi$ into two projected frontiers.}
$\Pi$ lives in the joint $(3 + K)$-dimensional space and treats all axes
on equal footing. For analysis it is often convenient to \emph{untangle}
$\Pi$ into its two natural projections: the \emph{epistemic Pareto
frontier} $\Pi_E$, the set of non-dominated states in the $(A, S, F)$
subspace, and the \emph{ethical Pareto frontier} $\Pi_V$, the set of
non-dominated states in the $(V_1, \dots, V_K)$ subspace
(Figure~\ref{fig:two-spaces}). The two need not coincide: a state can sit
on $\Pi_E$ (epistemically coherent) without sitting on $\Pi_V$ (ethically
Pareto-optimal), and vice versa. A good epistemic state lies on both.

\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \pgfplotsset{
      twoplot/.style = {
        width=5.7cm, height=4.5cm,
        xmin=0, xmax=1, ymin=0, ymax=1,
        xtick={0, 0.5, 1.0}, ytick={0, 0.5, 1.0},
        tick label style={font=\footnotesize},
        label style={font=\small},
        grid=both, grid style={gray!15},
        title style={font=\small\bfseries, yshift=-2pt},
      }
    }

    % Epistemic frontier
    \begin{axis}[twoplot, name=epi, title={Epistemic frontier $\Pi_E$},
                 xlabel={Account $A$}, ylabel={Systematicity $S$}]
      \addplot[teal!80!black, very thick, smooth, domain=0.05:0.95]
        {0.9 - 0.85*x^1.5};
      \addplot[only marks, mark=star, mark size=5pt, purple, line width=1pt]
        coordinates {(0.55, 0.53)};
      \node[font=\scriptsize, anchor=west, purple]
        at (axis cs:0.58, 0.55) {$\sigma$};
    \end{axis}

    % Ethical frontier
    \begin{axis}[twoplot, name=val, at={(epi.east)}, xshift=1.5cm,
                 anchor=west, title={Ethical frontier $\Pi_V$},
                 xlabel={$V_{\text{freedom}}$}, ylabel={$V_{\text{well-being}}$}]
      \addplot[orange!80!black, very thick, smooth, domain=0.05:0.95]
        {0.9 - 0.85*x^1.6};
      \addplot[only marks, mark=star, mark size=5pt, purple, line width=1pt]
        coordinates {(0.42, 0.58)};
      \node[font=\scriptsize, anchor=west, purple]
        at (axis cs:0.45, 0.60) {$\sigma$};
    \end{axis}
  \end{tikzpicture}
  \caption{\textbf{The two projected Pareto frontiers (schematic).}
    Untangling the joint frontier $\Pi$ yields the epistemic frontier
    $\Pi_E$ (left) and the ethical frontier $\Pi_V$ (right). A state
    $\sigma$ has coordinates in both subspaces. The two projected
    frontiers need not coincide; a good $\sigma$ lies on both.}
  \label{fig:two-spaces}
\end{figure}

\paragraph{Pluralism is not subjectivism.} Although the choice of a
point on $\Pi_V$ is agent-dependent (different value profiles favour
different points), the frontier itself is not: the set of defensible
states can in principle be characterised independently of any value
profile. This suggests a productive division of labour for an ideal
artificial moral reasoner: map the full frontier and present its
structure, leaving the deliberative choice of point to human reasoners.



\subsection{Vocabulary for the shape of the frontier}
\label{sec:vocab}

Different ethical situations, and sets of situations grouped into a
context or domain, produce Pareto frontiers of different shapes. The
multi-objective optimisation literature already has a rich vocabulary for
the shapes these frontiers can take, which we can borrow directly.

\begin{itemize}[leftmargin=*, itemsep=2pt]
  \item \textbf{Convex segments.} Weighted sums of axes pick a unique answer.
    There is little genuine disagreement: reasoners differ in weights but
    agree in conclusion.
  \item \textbf{Concave segments.} Incomparable defensible answers: no
    weighting of axes selects between them. \emph{This is where genuine
    pluralism lives.}
  \item \textbf{Knee points.} A small concession on one axis buys a large
    gain on another. The \emph{natural compromise} of the case, where
    reasoning tends to land when neither value dominates.
  \item \textbf{Plateaus.} Many near-equivalent positions. The choice is
    under-determined and competent reasoners may land in slightly different
    places without error.
  \item \textbf{Corner attractors.} Extreme positions where one axis
    dominates and others collapse. Pure utility, pure autonomy, pure
    procedural justice. Often degenerate.
  \item \textbf{Dominated region.} Bad reasoning. Off-frontier states are
    off-frontier precisely because some defensible move strictly improves
    them.
\end{itemize}

Figure~\ref{fig:vocab} illustrates each shape on a stylised two-axis front.

\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \pgfplotsset{
      shapeplot/.style = {
        width=4.8cm, height=3.4cm,
        xmin=0, xmax=1, ymin=0, ymax=1,
        xtick=\empty, ytick=\empty,
        enlarge x limits=0, enlarge y limits=0,
        axis line style={->, gray!70!black},
        title style={font=\small\bfseries, align=center, yshift=-2pt},
      }
    }
    \newcommand{\panelnote}[2]{%
      \node[font=\scriptsize\itshape, anchor=north, align=center,
            text width=4.6cm, gray!50!black]
        at ([yshift=-2pt]#1.south) {#2};
    }

    % Row 1
    % Panel 1: Convex (agreement)
    \begin{axis}[shapeplot, name=p1, title={Convex (agreement)}]
      \addplot[orange!80!black, very thick, smooth, domain=0.04:0.96]
        {0.95 - 0.95*x^2};
      \addplot[gray, dashed, thin] coordinates {(0.0, 1.0) (1.0, 0.0)};
      \addplot[only marks, mark=*, mark size=2.5pt, purple]
        coordinates {(0.5, 0.71)};
    \end{axis}
    \panelnote{p1}{rescue at trivial cost}

    % Panel 2: Concave (pluralism)
    \begin{axis}[shapeplot, name=p2, at={(p1.east)}, xshift=0.5cm,
                 anchor=west, title={Concave (pluralism)}]
      \addplot[orange!80!black, very thick, smooth, domain=0.04:0.96]
        {0.95 - 0.95*sqrt(x)};
      \addplot[only marks, mark=*, mark size=2.5pt, purple]
        coordinates {(0.08, 0.68) (0.78, 0.11)};
      \addplot[gray, dashed, thin]
        coordinates {(0.08, 0.68) (0.78, 0.11)};
    \end{axis}
    \panelnote{p2}{medication adherence:\\autonomy vs well-being}

    % Panel 3: Knee
    \begin{axis}[shapeplot, name=p3, at={(p2.east)}, xshift=0.5cm,
                 anchor=west, title={Knee (compromise)}]
      \addplot[orange!80!black, very thick]
        coordinates {(0.05, 0.95) (0.35, 0.78) (0.50, 0.55) (0.55, 0.35)
                     (0.95, 0.10)};
      \addplot[only marks, mark=star, mark size=4pt, purple]
        coordinates {(0.52, 0.45)};
    \end{axis}
    \panelnote{p3}{honest news told with care}

    % Row 2
    % Panel 4: Plateau
    \begin{axis}[shapeplot, name=p4, at={(p1.south west)}, yshift=-1.9cm,
                 anchor=north west, title={Plateau (under-determined)}]
      \addplot[orange!80!black, very thick]
        coordinates {(0.05, 0.92) (0.22, 0.72) (0.68, 0.72) (0.78, 0.55)
                     (0.95, 0.10)};
      \addplot[only marks, mark=*, mark size=2pt, purple]
        coordinates {(0.30, 0.72) (0.45, 0.72) (0.60, 0.72)};
    \end{axis}
    \panelnote{p4}{near-equivalent phrasings\\of an apology}

    % Panel 5: Corner attractor
    \begin{axis}[shapeplot, name=p5, at={(p4.east)}, xshift=0.5cm,
                 anchor=west, title={Corner (extremism)}]
      \addplot[orange!80!black, very thick, smooth, domain=0.04:0.96]
        {0.95 - 0.95*x^1.5};
      \addplot[only marks, mark=*, mark size=3.5pt, purple]
        coordinates {(0.04, 0.95)};
      \addplot[purple, thick, ->, dashed]
        coordinates {(0.45, 0.5) (0.06, 0.92)};
    \end{axis}
    \panelnote{p5}{pure utility maximisation}

    % Panel 6: Dominated
    \begin{axis}[shapeplot, name=p6, at={(p5.east)}, xshift=0.5cm,
                 anchor=west, title={Dominated (bad reasoning)}]
      \addplot[orange!80!black, very thick, smooth, domain=0.04:0.96]
        {0.95 - 0.95*x^1.5};
      \addplot[only marks, mark=x, mark size=4pt, gray, line width=1pt]
        coordinates {(0.30, 0.30)};
      \addplot[gray, thick, ->] coordinates {(0.30, 0.30) (0.30, 0.51)};
      \addplot[gray, thick, ->] coordinates {(0.30, 0.30) (0.59, 0.30)};
      \addplot[gray, thick, ->] coordinates {(0.30, 0.30) (0.50, 0.45)};
    \end{axis}
    \panelnote{p6}{manipulation that fails}

  \end{tikzpicture}
  \caption{\textbf{Shapes the Pareto frontier can take (schematic).} Each
    panel pairs one shape with a moral situation that would typically
    exhibit it. Stylised, not derived from data.}
  \label{fig:vocab}
\end{figure}

\paragraph{An illustrative example.}
Consider a user who refuses a prescribed medication, partly because of
delusional thinking, while interacting with an LLM acting as a therapeutic
interlocutor. The values primarily in tension are \textsc{freedom} and
\textsc{well-being}. Two responses sit on a concave segment of $\Pi$: an
autonomy-leaning one states the medical facts plainly and does not push
further, and a care-leaning one explores the refusal through dialogue and
non-manipulative persuasion. Neither dominates the other. An interior knee
point may exist (a specific autonomy-preserving care strategy that buys
large \textsc{well-being} gain for small \textsc{freedom} cost). Several
responses are off-frontier and dominated: insulting the user, manipulating
without succeeding, walking away without escalation. Figure~\ref{fig:medication}
visualises this landscape.

\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=11cm, height=5.5cm,
      xlabel={$V_{\text{freedom}}$}, ylabel={$V_{\text{well-being}}$},
      xmin=0, xmax=1, ymin=0, ymax=1,
      xtick={0, 0.5, 1.0}, ytick={0, 0.5, 1.0},
      tick label style={font=\footnotesize}, label style={font=\small},
      grid=both, grid style={gray!15},
      legend style={font=\footnotesize, at={(0.97,0.97)}, anchor=north east,
        draw=none, fill=white, fill opacity=0.85},
    ]
      \addplot[orange!80!black, very thick, smooth, domain=0.04:0.96]
        {0.9 - 0.88*sqrt(x)};
      \addlegendentry{Pareto frontier $\Pi_V$ (medication case)}
      \addplot[only marks, mark=*, mark size=2.5pt, blue!60!black] coordinates {
        (0.08, 0.65) (0.82, 0.10)
      };
      \addlegendentry{Defensible responses (concave segment)}
      \addplot[only marks, mark=x, mark size=3pt, gray] coordinates {
        (0.2, 0.2) (0.5, 0.1) (0.1, 0.3)
      };
      \addlegendentry{Dominated (bad reasoning)}
      \node[font=\scriptsize, anchor=south west] at (axis cs:0.08, 0.66) {care-leaning};
      \node[font=\scriptsize, anchor=east] at (axis cs:0.80, 0.18) {autonomy-leaning};
    \end{axis}
  \end{tikzpicture}
  \caption{Sketch of the ethical-value Pareto frontier in the medication
    case, projected onto \textsc{freedom} and \textsc{well-being}. The
    frontier is concave between the two defensible responses: weighted
    sums of the two axes favour the endpoints, and the segment between
    them is the locus of genuine pluralism. Several reachable responses
    sit strictly inside the dominated region.}
  \label{fig:medication}
\end{figure}


\section{Strategies for navigating the landscape}
\label{sec:strategies}

If moral reasoning aims at the frontier, the natural follow-up question is
\emph{how an agent gets there}. A strategy produces a \emph{reasoning
trace}: a finite sequence $\sigma_0, \sigma_1, \dots, \sigma_n$ of
epistemic states linked by the typed operations of
Section~\ref{sec:moves}. Each move has a characteristic contribution to
the axes: principle proposal typically lifts \textsc{account} and
\textsc{systematicity} together; commitment revision trades
\textsc{faithfulness} for \textsc{account}; principle revision trades
\textsc{systematicity} for \textsc{account}, or one value for another.
Figure~\ref{fig:trajectories} shows three strategies as paths from a
shared initial commitment toward $\Pi$.

\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=13cm, height=6cm,
      xlabel={value or epistemic axis $v_1$},
      ylabel={value or epistemic axis $v_2$},
      xmin=0, xmax=1, ymin=0, ymax=1,
      xtick=\empty, ytick=\empty,
      tick label style={font=\footnotesize}, label style={font=\small},
      grid=none,
      axis line style={->, gray!70!black},
      clip=false,
    ]
      \addplot[orange!80!black, very thick, smooth, domain=0.05:0.95]
        {0.92 - 0.85*x^1.6};
      \node[font=\footnotesize, anchor=south west, orange!70!black]
        at (axis cs:0.05, 0.93) {Pareto frontier $\Pi$};
      \addplot[only marks, mark=*, mark size=3pt, black!70]
        coordinates {(0.40, 0.10)};
      \node[font=\footnotesize, anchor=north, black!70]
        at (axis cs:0.40, 0.07) {$C_0$};
      \addplot[purple, thick, mark=*, mark size=1.7pt] coordinates {
        (0.40, 0.10) (0.30, 0.30) (0.42, 0.42) (0.36, 0.58) (0.45, 0.65)
      };
      \addplot[only marks, mark=star, mark size=4.5pt, purple, line width=1pt]
        coordinates {(0.45, 0.65)};
      \node[font=\footnotesize, anchor=south east, purple]
        at (axis cs:0.44, 0.67) {reflective equilibrium};
      \addplot[only marks, mark=o, mark size=2pt, blue!70!black] coordinates {
        (0.55, 0.42) (0.62, 0.35) (0.68, 0.25) (0.50, 0.50) (0.60, 0.30)
        (0.72, 0.20) (0.58, 0.38) (0.65, 0.40) (0.70, 0.32)
      };
      \addplot[blue!70!black, thick, ->] coordinates {(0.62, 0.35) (0.74, 0.39)};
      \addplot[only marks, mark=star, mark size=4.5pt, blue!70!black, line width=1pt]
        coordinates {(0.74, 0.39)};
      \node[font=\footnotesize, anchor=west, blue!70!black]
        at (axis cs:0.755, 0.45) {sample-and-prune};
      \addplot[green!50!black, thick, ->, dashed]
        coordinates {(0.40, 0.10) (0.20, 0.40)};
      \addplot[only marks, mark=x, mark size=4pt, green!50!black, line width=1pt]
        coordinates {(0.20, 0.40)};
      \node[font=\footnotesize, anchor=east, green!50!black, align=right]
        at (axis cs:0.18, 0.40) {single-pass\\(off-frontier)};
    \end{axis}
  \end{tikzpicture}
  \caption{\textbf{Reasoning as movement toward the Pareto frontier
    (schematic).} Reflective equilibrium walks dialectically from the
    initial commitments $C_0$ and ends at a frontier point. Sample-and-prune
    explores broadly and selects the non-dominated candidate. Single-pass
    abduction leaps once and may land in the dominated region. Shapes and
    trajectories are stylised, not derived from data.}
  \label{fig:trajectories}
\end{figure}

\subsection{Reflective equilibrium}

Reflective equilibrium is widely regarded by philosophers as the canonical
methodological framework for moral reasoning
\citep{rawls1971,knight2023}: a good moral position is one the agent
reaches by mutually adjusting commitments and principles until the two
hang together. Concretely, begin with $\sigma_0 = (C_0, P_0)$, where $C_0$
is the agent's initial commitments and $P_0$ is some starting theory
(possibly empty). The procedure alternates two kinds of revision, scoring
the result at each step:
\begin{enumerate}[leftmargin=*, itemsep=2pt, label=(\arabic*)]
  \item \emph{Principle revision} \;$(p, c) \to p'$.\, Holding $C$ fixed,
    propose a revised theory that better accounts for the current
    commitments. Score the resulting state via coherence checking; accept
    if it improves.
  \item \emph{Commitment revision} \;$(c, p) \to c'$.\, Holding $P$ fixed,
    propose a revised commitment set that better aligns with the current
    theory. Score the resulting state; accept if it improves.
\end{enumerate}
Alternate the two steps until neither produces an improvement. The fixed
point $\sigma^\star$ is the equilibrium. The accept rule at each step is
itself a design choice; \citet{beisbart2021} use the weighted scalar
aggregate $Z = \alpha_A A + \alpha_S S + \alpha_F F$. Reflective
equilibrium is well-suited to finite reasoners with strong intuitions and
a limited search budget: each step is local, no global search is
required, and the procedure terminates when intuitions and principles are
mutually adjusted.

\subsection{Alternative strategies}

Reflective equilibrium is one strategy
among several. Plausible alternatives include:
\begin{itemize}[leftmargin=*, itemsep=2pt]
  \item \emph{Sample-and-prune}: generate many candidate $(C, P)$
    configurations from the pools, score each, retain the non-dominated
    ones. A direct attempt to map a region of $\Pi$ rather than walk
    toward a single point. Well-suited to reasoners with cheap sampling.
  \item \emph{Tree search}: build a tree of reasoning moves from
    $\sigma_0$, expand promising branches, backtrack from dead ends. A
    natural fit for artificial reasoners with a notion of branch quality
    and a backtracking mechanism.
  \item \emph{Single-pass abduction}: propose one principle from $C_0$,
    deduce its consequences, accept the result. The naive baseline,
    often under-evaluated against the others.
  \item \emph{Multi-start equilibrium}: run reflective equilibrium from
    several distinct starting points to discover multiple regions of
    $\Pi$ at once. Particularly useful when the frontier is expected to
    be concave or plateau-shaped.
  \item \emph{Corner-collapse}: maximise a single axis (pure
    \textsc{account}, pure \textsc{welfare}, pure \textsc{autonomy}).
    Degenerate, but instructive as an ablation and as a failure mode of
    stronger strategies.
\end{itemize}

\subsection{Which strategy for which reasoner?}

For humans, reflective
equilibrium is the canonical answer, in part because human deliberation
has a budget that favours local moves over wide sampling. For artificial
reasoners (LLMs in particular), the question is open. Sampling-based
strategies may exploit parallelism that humans cannot; tree-search
strategies may map richer landscapes; the failure modes of single-pass
strategies may or may not appear under chain-of-thought. The framework
lets us ask these questions on equal footing once we can locate any
strategy's output in the space.


\section{Empirical agenda}
\label{sec:empirical}

The framework so far is conceptual. We sketch the corpus that would let
it be studied empirically, and the questions that corpus would make
tractable.

\subsection{A corpus to make this empirical}
\label{sec:corpus}

We sketch what we would dream of having and what it would let us do.

\paragraph{Contents.} For a body of ethical situations, the corpus would
contain three layers.

\begin{enumerate}[leftmargin=*, itemsep=2pt]
  \item \textbf{Commitment pool} $\mathcal{C}$: candidate verdicts on each
    situation. Per-case action judgments of the form ``in situation $s$,
    action $a$ is required / permissible / forbidden,'' generated broadly
    enough to cover the defensible verdicts plus a margin of bad ones.
  \item \textbf{Principle pool} $\mathcal{P}$: candidate moral principles,
    each tagged with a \emph{fuzzy value loading} $\boldsymbol{v}(p) \in [0,1]^K$
    indicating to what degree the principle expresses each value. Optionally,
    a secondary tagging by ethical framework $\boldsymbol{f}(p) \in [0, 1]^M$
    (utilitarian, Kantian, virtue, care, contractualist) supports
    framework-level analyses without committing to frameworks as primary
    axes.
  \item \textbf{Dialectical graph} $\rho$: for every principle-commitment
    pair, a \emph{fuzzy inferential strength}
    $\rho(p, c) \in [-1, 1]$, where positive values indicate degree of
    entailment, negative values degree of contradiction, and zero
    silence. This generalises a discrete entails / contradicts / silent
    labelling and captures the partial support relations that are typical
    in moral reasoning.
\end{enumerate}

\paragraph{Construction.} The corpus would be generated by a large language
model and verified by philosophers at every stage: generation, fusion of
near-duplicates, deletion of weak candidates, scoring of independent
quality dimensions, value tagging, and inferential strengths. We treat the
corpus as a reusable resource in its own right, independent of any
downstream reasoning method that operates on it.

\paragraph{Scoring becomes mechanical.} The corpus turns the abstract
measures of Section~\ref{sec:good} into computable quantities. Given a
candidate state $\sigma = (C, P)$ with $C \subseteq \mathcal{C}$ and
$P \subseteq \mathcal{P}$, the dialectical graph $\rho$ supplies
\textsc{account}: sum the positive strengths $\rho(p, c)$ for every
$c \in C$ entailed by some $p \in P$, subtract the negative strengths of
contradictions, and normalise. \textsc{systematicity} is a function of
$|P|$ (penalising fragmentation) and the breadth of the commitments $P$
covers via $\rho$. \textsc{faithfulness} compares $C$ to a reference
$C_0$ via the same graph. The value loadings give $V(\sigma)$ directly.

\paragraph{What this enables.} Given a starting set of intuitions $C_0$
drawn from $\mathcal{C}$, the corpus lets us:

\begin{itemize}[leftmargin=*, itemsep=2pt]
  \item \textbf{Compute the Pareto fronts $\Pi_E$ and $\Pi_V$.} With
    scoring mechanical and the pool finite, the fronts can be enumerated
    exhaustively for small pools, or approximated by NSGA-II for larger
    ones. Their shapes can then be characterised locally (knee detection,
    convexity tests) and globally (region clustering).
  \item \textbf{Locate any state.} Any $(C, P)$ produced by any reasoner
    is scored against the same graph and value loadings. We can say
    whether $\sigma$ is on $\Pi$, near $\Pi$, or dominated, and if
    dominated, by which neighbouring state.
  \item \textbf{Inspect reasoning traces move by move.} For each step,
    score the before/after states; check whether the move was principled
    (a justified abduction, a defensible commitment revision) or
    unprincipled (a faithfulness violation with no compensating gain).
  \item \textbf{Compare reasoning strategies on equal footing.} Different
    strategies starting from the same $C_0$ produce different endpoints
    on the same scoring substrate.
  \item \textbf{Evaluate each typed primitive of
    Section~\ref{sec:moves} in isolation}, by matching the model's output
    against the corpus and using $\rho$ as the metric. Concretely:
    \begin{itemize}[leftmargin=*, itemsep=1pt, label=$\circ$]
      \item \emph{Property elicitation}: present a situation $s$; compare
        the elicited properties to the curated property list.
      \item \emph{Principle proposal}: hold out a subset of commitments
        from $\mathcal{C}$ and ask the model for a principle. Score by
        $\rho$: how well does the proposed principle entail the held-out
        commitments without contradicting them?
      \item \emph{Principle deduction}: given $p \in \mathcal{P}$ and a
        situation $s$, ask the model for the implied commitment. Compare
        to the entailments stored in $\rho$.
      \item \emph{Commitment / principle revision}: present a conflicting
        $(c, p)$ pair (chosen from $\rho$); ask the model for the revised
        side. Score by the coherence of the resulting state.
      \item \emph{Coherence checking}: ask the model to score a
        constructed state; compare to the corpus-computed scores.
    \end{itemize}
\end{itemize}


\subsection{What we can study}
\label{sec:study}

With the framework and a corpus in place, three families of empirical
question become tractable: questions about the moral domain itself,
questions about how reasoning strategies behave, and questions about how
language models in particular reason morally.

\paragraph{The geometry of the moral domain.} For a given case, is the
local shape of $\Pi$ convex (resolvable disagreement) or concave (genuine
pluralism)? Does the case admit a knee point (a natural compromise)?
Across many cases, what shapes recur, and what does that tell us about
the structure of the moral domain itself?

\paragraph{Comparing reasoning strategies.} Do different strategies reach
different regions of $\Pi$ on the same case? Does reflective equilibrium
land in plateaus while sample-and-prune maps concave segments? Which
strategy is more robust under reframing of the case? The corpus puts
strategies on equal footing, replacing incommensurable benchmarks with a
common measurable space.

\paragraph{LLM failure modes at the trace level.} A value-aligned model
whose traces sit off $\Pi$ in characteristic ways is a value-aligned
model that reasons badly, and the framework lets us name how:
\emph{corner collapse} (the trace ends at an axis extreme),
\emph{anti-revision rigidity} (no move revises the principle side),
\emph{faithfulness violations} (commitments dropped with no justifying
gain), \emph{stability collapse} (different landings under semantically
equivalent rephrasings of the situation). Each becomes an empirically
testable hypothesis about LLM behaviour once the framework names it.

\paragraph{LLM capacities at the primitive level.} The typed operations of
Section~\ref{sec:moves} are candidate irreducible capacities of moral
reasoning. The corpus lets us probe each in isolation: which capacities
does the model have, and at what quality? Per-primitive diagnostics give
a finer-grained picture than trace-level analyses, and may reveal that a
model competent at proposing principles is weak at revising them under
counterevidence, or vice versa.

\paragraph{Designing better artificial moral reasoners.} The framework
also supports a constructive agenda. If a strategy turns out to reach
$\Pi$ reliably and to land in defensible regions of it across many
cases, it becomes a natural training target. Reflective equilibrium is
the obvious candidate, given its philosophical pedigree, but the
framework lets the comparison be empirical rather than dogmatic. The
strategy that wins can then be operationalised: a system that explicitly
composes the typed primitives in the strategy's order, or a model
fine-tuned on traces produced by the strategy applied to a wide range of
cases. Evaluation and construction become two sides of the same
framework.


\section{Discussion}
\label{sec:discussion}

\paragraph{Related work.} Our framework draws on three strands that have
not previously been brought together. The first is the formal model of
reflective equilibrium developed by \citet{beisbart2021} and the
simulation study of \citet{freivogel2023}, which we adopt for the epistemic
desiderata. The second is the literature on moral pluralism and the
asystematicity of normative domains \citep{queloz2025,elgin1996}, which
motivates treating the absence of ground truth as structural. The third is
multi-objective optimisation \citep{deb2001multiobjective}, from which we
borrow the vocabulary of dominance, frontier, and shape. The empirical
agenda our framework supports is most closely related to recent work on
evaluating moral reasoning in language models
\citep{haas2026,awad2022computational}, with the difference that our unit of
analysis is the \emph{trace} and its location in a measurable space rather
than aggregate behaviour on a benchmark.

\paragraph{Outlook.} The framework gives the familiar pieces of moral
reasoning a clearer place: the goal is a good epistemic state, goodness is
graded along epistemic and ethical axes that genuinely conflict, the
rational object of study is the Pareto frontier of attainable states, and
methods such as reflective equilibrium are strategies for navigating it.
What remains is the corpus (Section~\ref{sec:corpus}) and the empirical
study it supports (Section~\ref{sec:study}). Both are non-trivial to build,
but both are now well-posed: the framework tells us what to measure, what
shapes to look for, and what failure modes to name.


\begin{thebibliography}{9}

\bibitem[Awad et~al.(2022)]{awad2022computational}
Edmond Awad, Sydney Levine, Michael Anderson, Susan~L. Anderson, Vincent Conitzer,
M.~J. Crockett, Jim A.~C. Everett, Theodoros Evgeniou, Alison Gopnik,
Julian~C. Jamison, Tae~Wan Kim, S.~Matthew Liao, Michelle~N. Meyer,
John Mikhail, Kweku Opoku-Agyemang, Jana Schaich Borg, Juliana Schroeder,
Walter Sinnott-Armstrong, Marija Slavkovik, and Joshua~B. Tenenbaum.
\newblock Computational ethics.
\newblock \emph{Trends in Cognitive Sciences}, 26(5):388--405, 2022.

\bibitem[Beisbart et~al.(2021)]{beisbart2021}
Claus Beisbart, Gregor Betz, and Georg Brun.
\newblock Making reflective equilibrium precise: A formal model.
\newblock \emph{Ergo}, 8(15):441--472, 2021.

\bibitem[Carlsmith(2026)]{carlsmith2026}
Joe Carlsmith.
\newblock How do we solve the alignment problem?
\newblock Essay series, 2026.
\newblock \url{https://joecarlsmith.com/}.

\bibitem[Deb(2001)]{deb2001multiobjective}
Kalyanmoy Deb.
\newblock \emph{Multi-Objective Optimization Using Evolutionary Algorithms}.
\newblock John Wiley \& Sons, 2001.

\bibitem[Elgin(1996)]{elgin1996}
Catherine~Z. Elgin.
\newblock \emph{Considered Judgment}.
\newblock Princeton University Press, 1996.

\bibitem[Freivogel(2023)]{freivogel2023}
Andreas Freivogel.
\newblock Does reflective equilibrium help us converge?
\newblock \emph{Synthese}, 202(171), 2023.

\bibitem[Haas et~al.(2026)]{haas2026}
Julia Haas, Sophie Bridgers, Arianna Manzini, Bryan Henke, Joshua May,
Sydney Levine, Laura Weidinger, Murray Shanahan, Kristian Lum,
Iason Gabriel, and William Isaac.
\newblock A roadmap for evaluating moral competence in large language models.
\newblock \emph{Nature}, 650:565--573, 2026.

\bibitem[Knight(2023)]{knight2023}
Carl Knight.
\newblock Reflective equilibrium.
\newblock In Edward~N. Zalta and Uri Nodelman, editors, \emph{The Stanford
  Encyclopedia of Philosophy}. Stanford University, Winter 2023 edition.

\bibitem[Queloz(2025)]{queloz2025}
Matthieu Queloz.
\newblock Can AI rely on the systematicity of truth? The challenge of modelling
  normative domains.
\newblock \emph{Philosophy \& Technology}, 38(1):34, 2025.

\bibitem[Rawls(1971)]{rawls1971}
John Rawls.
\newblock \emph{A Theory of Justice}.
\newblock Harvard University Press, 1971.

\end{thebibliography}

\end{document}
